---
title: "data_8_acineto_abundance.Rmd"
author: "Daphne Garcia"
date: "2026-3-23"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
---

## Goals

I want to create an abundance heatmap for each metagenomic sample and
different species of Acineto:

1)  I will create a custom Kaiju database (copied from
    data_6_updated_snippy.Rmd). I will add Rath, Bois, Baret, Apis to
    this database, and update nodes.dmp

2)  Run 13.1 and 13.2

3)  in bash, write chunk that will make a .tsv or .csv: header, sample,
    pol, nect,baret, bois, rath, apis. total classified, total acineto
    (all pol, nect, etc.)

4)  in R, open .tsv and make a heatmap in base R

| Number | Acinetobacter species |
|--------|-----------------------|
| 1      | Pollinis              |
| 2      | Nectaris              |
| 3      | Baretiae              |
| 4      | Rathkeae              |
| 5      | Boissieri             |
| 6      | Apis                  |

### 13 Kaiju CUSTOM database

I will modify scripts from data_5_SNP which use a custom Kaiju database
and split contigs into either pollinis or nectaris. From here, I will
feed this pipeline back into 14. Snippy

#### 13.0 create a custom kaiju database

I copied the folder data_5/9_kaiju/9.0 Acineto ref kaiju db:

it contains .faa I also made fake names.dmp and nodes.dmp by copying
A-pol and A-nect species info from node/name.dpm files in
daphne/floral/databases/kaiju-nr-euk (this file is very big so I opened
it with VScode)

```{bash kaiju, eval = FALSE}
  
# I ran this from inside data_5/9/9.0  
cd /local/workdir/Daphne/Floral_metagenomics/data_8_acineto_abundance/13_Kaiju/13.0_Acineto_ref_kaijudb


# I added prefixes to all references to match NCBI nodes, found in names.dmp (manually change file
# names in the code)
awk '/^>/ {print ">2605383"; next} {print}' A-baretiae-B10A.faa > A-baretiae-B10A_protein.kaiju.faa
  
#Then I added them all to one doc:
"" > all_ref_acineto_protein_genomes.faa  ## clear this file
for file in ./*kaiju.faa; do 
  cat $file >> all_ref_acineto_protein_genomes.faa
done

#now, make the kaiju database .fmi
export PATH=/programs/kaiju-1.10.0/bin:$PATH
kaiju-mkbwt -n 5 -a ACDEFGHIKLMNPQRSTVWY -o all_ref_acineto_protein_genomes all_ref_acineto_protein_genomes.faa
kaiju-mkfmi all_ref_acineto_protein_genomes
# don't forget to delete .bwt and .sa files

```

the file `all_ref_acineto_protein_genomes.fmi` is now created in
`data_5/9/9.0`

##### DO ONCE: move bowtie host filtered reads into data_8:

copy bowtie file names into data_8_acineto_abundance

```{bash}

cd /local/workdir/Daphne/Floral_metagenomics/data_8_acineto_abundance/3_bowtie

for sample in ../../data_3/03_bowtie/2_bowtie_output_sam/*; do
  base=$(basename "$sample"); base=${base#*_}
  mkdir $base
  cp ${sample}/SAMPLE_host_removed_R1.fastq $base
  cp ${sample}/SAMPLE_host_removed_R2.fastq $base
done 
#took 1 minute

```

#### 13.1 run kaiju on all the files

NOTE: if multiple tmux are running kaiju, it will crash. just run one

```{bash tmux_kaiju, eval = FALSE}

cd /local/workdir/Daphne/Floral_metagenomics/data_8_acineto_abundance
export PATH=/programs/kaiju-1.10.0/bin:$PATH

  for sample in 3_bowtie/*; do
    R1="${sample}/SAMPLE_host_removed_R1.fastq"
    
    echo $(date)
    export PATH=/programs/kaiju-1.10.0/bin:$PATH
    R2="${R1/R1.fastq/R2.fastq}"
    output_name=$(echo "$R1" | cut -d'/' -f2)
    kaiju_path="/local/workdir/Daphne/Floral_metagenomics/data_8_acineto_abundance/13_Kaiju/13.0_Acineto_ref_kaijudb"
    
    kaiju -t "${kaiju_path}/nodes.dmp" \
    -f "${kaiju_path}/all_ref_acineto_protein_genomes.fmi" \
    -i $R1 -j $R2 \
    -o 13_Kaiju/1_raw_kaiju/${output_name}_kaiju.txt \
    -z 30
    
  done
# Takes 10 minutes

```

Finished in 9.5 minutes

##### all_classified_kaiju_stats.out

Now this script creates `all_classified_kaiju_stats.out`, to check check
percent of lines that are "classified vs unclassified Output: number of
classified reads, total reads, and percent, for each sample

```{bash}

for file in 1_raw_kaiju/*kaiju.txt; do
  echo $file >>  1_raw_kaiju/classified_vs_unclassified_stats.out
   #num files with C, total files, C's percent of total
  awk '{total++} $1 == "C" {c++} END {print c, total, c/total}'  $file >>  1_raw_kaiju/classified_vs_unclassified_stats.out
done
# takes 30 sec
```

Outputs for both of these scripts are saved in
data_8/13_Kaiju/1_raw_kaiju

When comparing old (Mar 23 data_6_updated) and new (Mar 23
data_8_acineto_abd), the number of classified reads per sampe stays
relatively the same, but the percent changes.This is because data_6 used
reads from data_updated, which used the bowtie host filtering
incorrectly. data_8 uses reads from data_3, which removes more reads,
and increases the percent of classified reads.

| Sample name    | classified | unclassified | percent   |
|----------------|------------|--------------|-----------|
| 44/FN118 old   | 61959      | 1402578      | 0.0441751 |
| 44/FN118 new   | 61442      | 1008371      | 0.0609319 |
|                |            |              |           |
| 45/FN128 old   | 70063      | 1772434      | 0.0395293 |
| 45/FN128 new   | 69930      | 201663       | 0.346767  |
|                |            |              |           |
| 48/22FN018 old | 68885      | 899316       | 0.0765971 |
| 48/22FN018 new | 68122      | 575121       | 0.118448  |

#### 13.2 assign genus taxonomy to kaiju reads

Once kaiju files are made, isolate the acinetobacter files. Run
kaiju-addTaxonNames -r genus to add genus taxonomy to all reads

```{bash}
  
cd /local/workdir/Daphne/Floral_metagenomics/data_8_acineto_abundance/13_Kaiju/1_raw_kaiju 

kaiju_path="/local/workdir/Daphne/Floral_metagenomics/data_8_acineto_abundance/13_Kaiju/13.0_Acineto_ref_kaijudb"

for files in ./*.txt; do
  name=$(basename "$files" .txt)_species.out
  
  kaiju-addTaxonNames -t ${kaiju_path}/nodes.dmp \
  -n ${kaiju_path}/names.dmp \
  -i $files \
  -r species \
  -o ../2_kaiju_taxonomy/$name
done
#Took 10 seconds  
```

Outputs for these scripts are saved in
`data_6/13_kaiju/2_kaiju_taxonomy`

For an output file with all sample's stats, run this script: print to
acineto_stats.out number of "Acinetobacter" per sample

```{bash}
cd /local/workdir/Daphne/Floral_metagenomics/data_8_acineto_abundance/13_Kaiju/2_kaiju_taxonomy

for files in ./*_species.out; do
  echo $(basename "$files") >> acineto_stats.out
  awk '/pollinis/ {count++} END {print "pollinis = "count}' $files >> acineto_stats.out
  awk '/nectaris/ {count++} END {print "nectaris = "count}' $files >> acineto_stats.out
  awk '/apis/ {count++} END {print "apis = "count}' $files >> acineto_stats.out
  awk '/baretiae/ {count++} END {print "baretiae = "count}' $files >> acineto_stats.out
  awk '/rathckeae/ {count++} END {print "baretiae = "count}' $files >> acineto_stats.out
  awk '/boissieri/ {count++} END {print "boissieri = "count}' $files >> acineto_stats.out

  echo "" >> acineto_stats.out
done

# took 2 min
```

Output is `data_6/13_kaiju/2_kaiju_taxonomy/acineto_stats.out`

Comparing acineto_stats.out from before using the old kaiju_nr_euk db
and the new custom pol-nect database: we are getting around the same
amount of hits (a little more) with our new database.

OLD: 3/23 from data_6 (worse bowtie host filtering, only pol and nect)

NEW: 3/23 from data_8 (bowtie filtered reads from data_3, pol, nect,
apis, etc.)


FN128_kaiju_species.out
pollinis = 51935
nectaris = 3160

| Sample name    | pollinis | nectaris |
|----------------|----------|----------|
| 44/FN118 old   | 16211    | 9545     |
| 44/FN118 new   | 12681    | 5710     |
|                |          |          |
| 45/FN128 old   | 52775    | 3521     |
| 45/FN128 new   | 51935    | 3160     |
|                |          |          |
| 48/22FN018 old | 40361    | 5628     |
| 48/22FN018 new | 39671    | 3108     |

for 22FN018 and FN118: number of nectaris reads decreased because many of those reads are 
now classified as other species. In data_8/13.1, generally there were the same
number of classified reads as in data_6, and due to that, some of those reads
are better classified as apis or rathckeae and the total number of nectaris 
reads decreases


#### 3. making a dataframe with abundance csv from 13.2

Now that I have data_8/13.2/acineto_stats.out: I want to transform this data
into a .tsv or .csv format with columns:
```{bash}
cd /local/workdir/Daphne/Floral_metagenomics/data_8_acineto_abundance/13_Kaiju/2_kaiju_taxonomy

Outfile="../../14_heatmap/acineto_abundances_Mar23.csv"

# Add headers and clear existing data
echo "\"Samples\",\"A. pollinis\", \"A. nectaris\", \"A. apis\", \"A. baretiae\", \"A. rathckeae\", \"A. boissieri\", \"Total Classified\", \"Total Acinetobacter\", \"Total Reads\"" > $Outfile

for files in ./*_species.out; do

  # store all of the values for .csv in variables:
  name=$(basename "$files"); name=${name%%_*}
  pol=$(grep -c 'pollinis' $files)
  nect=$(grep -c 'nectaris' $files)
  apis=$(grep -c 'apis' $files)
  baret=$(grep -c 'baretiae' $files)
  rath=$(grep -c 'rathckeae' $files)
  bois=$(grep -c 'boissieri' $files)
  class=$(grep -Pc 'C\t' $files)
  total=$(expr $pol + $nect + $apis + $baret + $rath + $bois)
  readnum=$(wc -l < $files)
  
  # add all values in csv format (comma) to Outfile
  echo " \"${name}\", \"${pol}\",  \"${nect}\", \"${apis}\", \"${baret}\", \"${rath}\", \"${bois}\",  \"${class}\",  \"${total}\",  \"${readnum}\"" >> $Outfile
  
done

```

This creates a .csv file `data_8_acineto_abundance/14_heatmap/acineto_abundances_Mar23.csv`

#### 4. R: updating .csv to heatmap in R

First, load libraries, set wd() and load data table
```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)

setwd("/local/workdir/Daphne/Floral_metagenomics/data_8_acineto_abundance/14_heatmap")
abundance <- read_csv("acineto_abundances_Mar23.csv")

```

Second, filter out 22FN006, FN062, and FN140, because they don't have any classified reads 
```{r}

abundance <- abundance %>%
  filter(!Samples %in% c("22FN006", "FN062", "FN140"))
```

Third, scale by all classified, and scale by all acineto, then pivot longer

```{r}
#this contains acineto abundances scaled by all classified reads
scale_read_abd <- abundance %>%
  mutate(across(-c(Samples, `Total Classified`, `Total Acinetobacter`, `Total Reads`),
                ~ . / `Total Reads`)) %>%
  select(-`Total Classified`, -`Total Acinetobacter`, -`Total Reads`) %>%
  pivot_longer(cols=c('A. pollinis', 'A. nectaris', 
                      'A. apis', 'A. baretiae', 
                      'A. rathckeae', 'A. boissieri'),
                    names_to='Species',
                    values_to='Relative_Abundance') 
  
  
  
#this contains acineto abundances scaled by total acineto reads  
scale_acineto_abd <- abundance %>%
  mutate(across(-c(Samples, `Total Classified`, `Total Acinetobacter`, `Total Reads`),
                ~ . / `Total Acinetobacter`)) %>%
  select(-`Total Classified`, -`Total Acinetobacter`) %>%
  pivot_longer(cols=c('A. pollinis', 'A. nectaris', 
                      'A. apis', 'A. baretiae', 
                      'A. rathckeae', 'A. boissieri'),
                    names_to='Species',
                    values_to='Relative_Abundance') 


```

fourth: make a heatmap
```{r}



ggplot(scale_read_abd, aes(Samples, Species, fill= Relative_Abundance)) + 
  geom_tile(color = "white", lwd = 1.5,linetype = 1) +
  geom_tile(aes(fill = ifelse(Relative_Abundance < 1e-4, NA, Relative_Abundance)),
                  color = "white", lwd = .5, linetype = 1) +
        scale_fill_gradientn(
          colours = c("#075AFF", "#FFFFCC", "#FF0000"),
          values  = scales::rescale(c(0.0025, 0.025, 0.25)),
          limits  = c(2.5e-4, 2.7e-1),
          na.value = "white") +
  labs(title="Relative abundance of species by sample (scaled by total number of reads)",
        x ="Sample", y = "species")+
  scale_x_discrete(position = "top") +
  scale_y_discrete(limits = rev(c("A. pollinis", "A. nectaris", "A. apis",
                                 "A. baretiae", "A. rathckeae", "A. boissieri"))) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0),
        axis.text.y = element_text(color = "grey20", size = 20, angle = 0, hjust = 1, vjust = 0, face = "plain"),  
        axis.title.x = element_text(color = "grey20", size = 12, angle = 0, hjust = .5, vjust = 0, face = "plain"),
        title = element_text(size = 20),
        axis.title.y = element_text(color = "grey20", size = 12, angle = 90, hjust = .5, vjust = .5, face = "plain"))
    



```


####Counting number of reads per file
I need this for the supplemental data S1
```{bash}
cd /local/workdir/Daphne/Floral_metagenomics/data_8_acineto_abundance/3_bowtie

"" > all_read_counts.txt  
for sample in * ; do
  read_count=$(grep -c '^@.*/[0-9]$' "${sample}/SAMPLE_host_removed_R1.fastq")
  echo "${sample}  ${read_count}" >> all_read_counts.txt 
done
```



