---
title: 'data_7_barplot_NMDS.Rmd'
author: "Daphne"
date: "2026-03-04"
output: html_document
editor_options: 
  chunk_output_type: console
---


## Goals
Here I will create stacked barplots from the shotgun metagenomic data
I will use kaiju to classify reads, copied from data_3/09_kaiju
and create a taxonomy and OTU table
input these tables into phyloseq in R
Create a nested stacked barplot with fantaxtic


### Copying kaiju files
```{bash}

cd /local/workdir/Daphne/Floral_metagenomics/data_7_barplot_NMDS
mkdir 9_kaiju
cp -r  ../data_3/09_kaiju/1_raw_kaiju_euk/ 9_kaiju/
## I put this directory inside 

```

Using kaiju2table, I will create a .tsv file that contains the number and 
percent of reads per sample that match to a specific genus. The percent and 
taxon id will be used to make an otu table for phyloseq, and the taxon id and 
taxon names will be used to make the taxonomy table for phyloseq

### OTU and taxonomy table kaiju2table
```{bash}
cd /local/workdir/Daphne/Floral_metagenomics/data_7_barplot_NMDS/9_kaiju/1_raw_kaiju_euk
export PATH=/programs/kaiju-1.10.0/bin:$PATH
kaiju_path="/local/workdir/Daphne/Floral_metagenomics/databases/kaiju_nr_euk"

kaiju2table \
-t "${kaiju_path}/nodes.dmp" \
-n "${kaiju_path}/names.dmp" \
-u \
-r genus \
-l superkingdom,phylum,class,order,family,genus \
-o ../OTU_and_taxonomy_table.tsv \
1_22FN006_kaiju.txt 2_22FN061_kaiju.txt 3_22FN096_kaiju.txt 4_FN106_kaiju.txt 1_22FN008_kaiju.txt 2_22FN062_kaiju.txt 3_22FN098_kaiju.txt 4_FN116_kaiju.txt 1_22FN018_kaiju.txt 2_22FN063_kaiju.txt 3_22FN100_kaiju.txt 4_FN118_kaiju.txt 1_22FN020_kaiju.txt 2_22FN064_kaiju.txt 3_22FN102_kaiju.txt 4_FN128_kaiju.txt 1_22FN021_kaiju.txt 2_22FN065_kaiju.txt 3_22FN103_kaiju.txt 4_FN131_kaiju.txt 1_22FN025_kaiju.txt 2_22FN066_kaiju.txt 3_FN011_kaiju.txt 4_FN136_kaiju.txt 1_22FN028_kaiju.txt 2_22FN068_kaiju.txt 3_FN021_kaiju.txt 4_FN140_kaiju.txt 1_22FN029_kaiju.txt 2_22FN070_kaiju.txt 3_FN026_kaiju.txt 4_FN142_kaiju.txt 1_22FN031_kaiju.txt 2_22FN071_kaiju.txt 3_FN030_kaiju.txt 4_FN144_kaiju.txt 1_22FN038_kaiju.txt 2_22FN072_kaiju.txt 3_FN051_kaiju.txt 4_FN150_kaiju.txt 1_22FN039_kaiju.txt 2_22FN073_kaiju.txt 3_FN053_kaiju.txt 4_FN153_kaiju.txt 1_22FN057_kaiju.txt 2_22FN081_kaiju.txt 3_FN054_kaiju.txt 4_FN174_kaiju.txt 1_22FN058_kaiju.txt 2_22FN082_kaiju.txt 3_FN062_kaiju.txt 4_FN176_kaiju.txt 1_22FN059_kaiju.txt 2_22FN083_kaiju.txt 4_FN075_kaiju.txt
  # takes 20 seconds

```

-t the node values, and how they are connected order --> family --> genus --> species
-n the names that correspond to node value
-u removes "unclassified" 
-m filter out taxa with under 10% abundances (REMOVED, bc doesn't work with -u)
-l all taxonomy groups (we're specifying  superkingdom,phylum,class,order,family,genus)
-r taxonomic level (genus)
-o output .tsv for the OTU abundance table and .out for the kaiju.txt classified/unclassified from data_3


## SWITCH TO R ##

### Tidyverse to reformat taxonomy and OTU/Abundance tables

Currently, the OTU/abundance table `genus_table.tsv` is formatted:
file                    percent        reads   taxon_id      taxon_name
1_22FN006_kaiju.txt     100.000000      16      NA      unclassified
2_22FN061_kaiju.txt     8.795682        929     44471   Microcoleus
2_22FN061_kaiju.txt     8.256012        872     13687   Sphingomonas

but should be:
taxon_id   sample1   sample2   sample3
44471      0         929       500
13687      0         872       120

```{r}
library(tidyverse)
setwd("/local/workdir/Daphne/Floral_metagenomics/data_7_barplot_NMDS")
otu_tax_table <- read_tsv("9_kaiju/OTU_and_taxonomy_table.tsv")

# remove "1_" and _kaiju.txt" from sample names
otu_tax_table <- otu_tax_table %>% 
  mutate(across('file', str_replace, '1_|2_|3_|4_', '')) %>%
  mutate(across('file', str_replace, '_kaiju.txt', ''))

otu <- otu_tax_table %>%
  filter(!is.na(taxon_id)) %>%      #remove "NA" (unclassified reads)
  select(file, taxon_id, reads) %>%  # just keep taxon_id and reads columns, and flip df sideways
  pivot_wider(names_from = file,
              values_from = reads,
              values_fill = 0)

write.table(otu, "10_phyloseq/otu_table.tsv", sep="\t", quote=FALSE, row.names=FALSE)

```



The taxonomy table `taxonomy_table.tsv` is formatted:
file                    percent        reads  taxon_id  taxon_name
1_22FN006_kaiju.txt     0.000000        0       10239   Viruses
2_22FN061_kaiju.txt     8.795682        929     44471 Bacteria;Cyanobacteriota;Cyanophyceae;Oscillatoriales;Microcoleaceae;Microcoleus;
2_22FN061_kaiju.txt     8.256012        872     13687  Bacteria;Pseudomonadota;Alphaproteobacteria;Sphingomonadales;Sphingomonadaceae;Sphingomonas;

But should be:
taxon_id kingdom phylum class order family genus
44471 Bacteria Cyanobacteriota Cyanophyceae Oscillatoriales Microcoleaceae Microcoleus
13687 Bacteria Pseudomonadota Alphaproteobacteria Sphingomonadales Sphingomonadaceae Sphingomonas

```{r}
otu_tax_table <- read_tsv("9_kaiju/OTU_and_taxonomy_table.tsv")

tax <- otu_tax_table %>%
  filter(!is.na(taxon_id)) %>%
  distinct(taxon_id, taxon_name) %>%
  separate(taxon_name,
           into=c("kingdom","phylum","class","order","family","genus"),
           sep=";",
           fill="right")

write.table(tax, "10_phyloseq/taxonomy_table.tsv", sep="\t", quote=FALSE, row.names=FALSE)
```
  note: there's a lot of plants and fungi in the taxonomy database, and therefore
  in our samples! Interesting


### Making sample_data and metadata by hand

I copied Vivi's metadata from the phyloseq object `physeq1146_flowers@sam_data`,
and inputted the missing samples (22FN031, 22FN083, 22FN102, FN062, FN140) from
my printed metagenomic metadata sheet. 
Then, I uploaded it onto Rstudio in data_7/10_phyloseq, and I will add it to the
phyloseq object

```{r}
sam_data <-read.csv("10_phyloseq/sample_metadata.csv", row.names = "row")

```


### merging tables into a phyloseq object

Taken and edited from Mar's class pdf 03_PreProcessing.html, in my downloaded files
```{r}
library(phyloseq)

kaiju_phyloseq <- 
  phyloseq(otu_table(as.matrix(otu), taxa_are_rows=TRUE),
         tax_table(as.matrix(tax)),
         sample_data(sam_data))

```


### stacked barplot code from Vivi
This is taken from Daphne/Floral_16S/Vivi_barplot_NMDS_analysis.Rmd

```{r}

# Transform to relative abundance
kaiju_physeq_relative <- kaiju_phyloseq %>%
  transform_sample_counts(function(x) x / sum(x))

# remove any samples (22FN006) that didn't have any genera
kaiju_physeq_relative <- transform_sample_counts(
  kaiju_phyloseq,
  function(x) {
    if(sum(x) == 0) x else x / sum(x)
  }
)

#restructuring taxonomy names to fix the barplot
colnames(tax_table(kaiju_phyloseq)) <- c(
  "kingdom",
  "phylum",
  "class",
  "order",
  "family",
  "genus"
)

# This is fantaxtic code, it creates a ps_obj that is used for further stacked
#barplot code. Change Order and Genus taxonomic levels here:
k_top_nested <- nested_top_taxa(kaiju_physeq_relative,
                              top_tax_level = "order",
                              nested_tax_level = "genus",
                              n_top_taxa = 7, 
                              n_nested_taxa = 7, 
                              nested_merged_label = "NA and other <tax>")

  #took 3 min

# This is following code from github tutorial. Adding title, colors, and legend
plot_nested_bar(k_top_nested$ps_obj,
                top_level = "order",
                nested_level = "genus",
                nested_merged_label = "NA and other <tax>",
                legend_title = "Order and Genus",
                merged_clr = "black") +
  facet_wrap(~SampleSource2,
             scales = "free_x") +
  labs(title = "METAGENOMIC Relative abundances of the top 6 genera for each of the top 3 Orders") +
  theme(plot.title = element_text(hjust = 0.5, 
                                  size = 10, 
                                  face = "bold"),
        legend.key.size = unit(10, 
                               "points"))

```


### making abundance tables to know which genera to include
these abundance tables will tell me which genera to include in fantaxtic
```{r}

# making table that gives top abundance per sample
genus_long <- psmelt(kaiju_physeq_relative)

top_genus_per_sample <- genus_long %>%
  group_by(Sample) %>%
  slice_max(Abundance, n = 1, with_ties = FALSE) %>%
  select(Sample, genus, Abundance) %>%
  rename(
    Most_abundant_genus = genus,
    Abundance = Abundance
  )

view(top_genus_per_sample)


#making table that gives all abundances over 10%

genera_over15 <- genus_long %>%
  filter(Abundance > 0.15)

library(tidyr)
genera_over15_table <- genera_over10 %>%
  select(genus, Sample, Abundance) %>%
  pivot_wider(
    names_from = Sample,
    values_from = Abundance,
    values_fill = 0
  )

view(genera_over15_table)

#Making list of number of samples that contain over 10% per genera
genus_sample_counts <- genera_over10_table %>%
  mutate(n_samples = rowSums(across(-genus, ~ . > 0))) %>%
  select(genus, n_samples)


```


### Fantaxtic plotting barplots
#### This is Sophia's code:


```{r}

remotes::install_github("mikemc/speedyseq")
library(phyloseq)

# create df of genus, technically not needed since fantaxtic transform sample counts
genus_df_abs <- kaiju_physeq_relative %>%
  psmelt() %>%
  mutate(genus = case_when(is.na(genus)~"Other",
                            TRUE ~ genus))

# I want these genera, which are > 15% abundant in at least one sample
custom_genus <- c(
                 "Acinetobacter",
                  "Actinobacillus",
                  "Alternaria",
                  "Arsenophonus",
                  "Aureobasidium",
                  "Bacillus",
                  "Buchnera",
                  "Erwinia",
                  "Exiguobacterium",
                  "Izhakiella",
                  "Labilibacter",
                  "Metschnikowia",
                  "Pantoea",
                  "Pseudomicrostroma",
                  "Pseudomonas",
                  "Puccinia",
                  "Rosenbergiella",
                  "Venturia")


# add column "top_genus": if "genus" within custom_genus df (above), then add "this_is_top_genus" to the "top_genus" column. Otherwise, add "Other" to this row.
abund_taxa_abs <- genus_df_abs %>%
  dplyr::mutate(
    "top_genus" = case_when(
      genus %in% custom_genus ~ "this_is_top_genus",
      TRUE ~ "Other"))

#Now we're doing taxa. add column "top_genus": if "genus" within custom_genus df (above), then add "this_is_top_genus" to "top_genus" column. Otherwise add "Other" to this row.
tax_tab <- as.data.frame(tax_table(kaiju_physeq_relative))
tax_tab$top_genus <- dplyr::case_when(
  tax_tab$genus %in% custom_genus ~ "this_is_top_genus",
  TRUE ~ "Other")

# redo ranking for fantaxtic
tax_tab <- tax_tab %>%
  dplyr::select(top_genus, kingdom, phylum,
                class, order, family, genus)

# add tax table to physeq
tax_table(kaiju_physeq_relative) <- tax_table(as.matrix(tax_tab))

# or lets subset phyloseq object to a subset 
ps_tmp <- subset_taxa(kaiju_physeq_relative, 
                      genus %in% custom_genus)

taxids <- taxa_names(ps_tmp)

ps_collapsed <- collapse_taxa(kaiju_physeq_relative, 
                              taxa_to_keep = taxids,
                              merged_label = "Other")

# here lets just follow the example
top_nested <- nested_top_taxa(ps_collapsed,
                              top_tax_level = "order",
                              nested_tax_level = "genus",
                              top_merged_label = "other",
                              n_top_taxa = 7, 
                              n_nested_taxa = 9,
                              include_na_taxa = T,
                              by_proportion = T)

plot_nested_bar(ps_obj = top_nested$ps_obj,
                top_level = "order",
                nested_level = "genus",
                top_merged_label = "other",
                merged_clr = "black",
                include_rank = T) +
  facet_wrap(~SampleSource2,
             scales = "free_x") +
  labs(title = "Relative abundances of genera with >15% abundance", subtitle = "Metagenomic data", y = "Relative Abundance") +
  theme(legend.position = "right",
        axis.title.x = element_blank(),
        axis.ticks.x = element_blank(),
        strip.text = element_text(size = 10, face = "bold"),
        axis.title.y = element_text(size = 10, face = "bold"),
        axis.text.y = element_text(size = 9))



```



