---
title: "metagenomic_workflow"
author: "Daphne Garcia"
date: "2025-06-23"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
---


## Goals
This is a second version of the original metagenomic_workflow.Rmd, with new updates:

- bowtie host removal includes Malus chloroplast, since that is a large part of dandelion bins
-metaspades AND megaHIT
- metablast2 with new parameters


## Steps

Previously done in metagenomic_workflow.Rmd:

1.  FastQC quality analysis
2.  Trimmomatic adapter trimming
3.  Bowtie2 host-read filtering 

(Found in local.workdir/Daphne/Floral_metagenomics/data)

+ Malus chloroplast 
3.5 Bowtie2 read filtering with mito
4.  MetaSpades assembly
5.  megaHIT assembly
5.  metaQuast quality assessment of both
6.  megabat2
6.1 megabat2 + parameters
7.1 checkM
6.2 megabat2 + parameter
7.2 checkM
6.3 megabat2 + parameter
7.3 checkM
  
# TASKS # 
bowtie build, 
bowtie align

- while bowtie is running: 
      go do metabat alterations on metagenomic_workflow.Rmd
- When bowtie is finished
      start metaspades
      run kraken on bowtie-filtered reads

 
 FOR REFERENCE: 
| Number | reference genome     |
|--------|----------------------|
| 1      | Cichorium intybus    |
| 2      | Malus domestica      |
| 3      | Lotus japonicus      |
| 4      | Ranunculus sardous   |
| 5      | Trifolium repens     |
| 6      | Matricaria discoidea |
| 7      | Alliaria petiolata   |
| 8      | Solidago caesia      |
| 9      | Lonicera japonica    |
| 10     | Linaria vulgaris     |
 
### 3. Bowtie2

We're gonna run the data/02_trimmed_paired_filtered reads and filter against Malus (apple), Taraxacum (dandelion), Lotus (trefoil), and Lonicera (honeysuckle) mitochondria, which contaminated many samples 

#### 3.1 Download/build reference indices

First, I need to concatenate all mitochondrial genomes since they're small, and build an index out of the concatenated fasta. 

Note: when I started, I ftp-downloaded all 4 mito genomes into /03_Flower_host_genomes/Mitochondria_indices

```{bash bowtie2_build_mito_reference_index, eval=FALSE}

#enter right directory and load bowtie2
cd /local/workdir/Daphne/Floral_metagenomics/data/03_Flower_host_genomes/Mitochondria_indices
export PATH=/programs/bowtie2-2.5.1-linux-x86_64:$PATH

#combine all 4 mitochondrial genomes
cat ./*.fasta > ./mito_genomes.fasta

#Index for Cichorium intybus (dandelion, chickory, hawkweed)
bowtie2-build ./mito_genomes.fasta mito_index

```

#### 3.2 Bowtie host filtering

Here we are in the data/ folder, looping through 
`data/02_trim_unzipped_and_host-grouped` R1 and R2 reads, and aligning them to 
the mitochondria reference genome index: 
`data/03_Flower_host_genomes/Mitochondria_indices`

```{bash bowtie2_loop, eval = FALSE}

#enter right directory and load bowtie2
cd /local/workdir/Daphne/Floral_metagenomics/data/02_trim_unzipped_and_host-grouped
export PATH=/programs/bowtie2-2.5.1-linux-x86_64:$PATH
 
for sample in ./* ; do
  cd $sample
  for R1 in ./*_R1.fastq; do
    base="$(basename "${R1%%_R1.fastq}" | cut -d'_' -f1-4)"
    output_dir="../../../data_updated/03_Bowtie2_host_filter/${sample}/${base}"
    mkdir $output_dir
    bowtie2 --sensitive-local -p 32 --seed 4 -x ../../03_Flower_host_genomes/Mitochondria_indices/mito_index -1 $R1 -2 ${R1/R1.fastq/R2.fastq} -S "$output_dir/SAMPLE_mapped_and_unmapped.sam";
  done     
  cd - > /dev/null
done   

#Faster way with tmux: in terminal, do "sample=1_Chichorium_samples", etc, and run only the nested for loop of the function
    
# --sensitive-local (not very-sensitive) because our ref host genomes are usually different genuses of the actual host
# -p is the number of parallel threads 
#--seed 4 is a random number to make it so that work is reproducible 
# -x The basename of the index for the reference genome variable $index
# -1 and -2 are the paired inputs R1 and R2
# -S Summary of index settings and input names/lengths    
 
```

Outputs are `paired_1370.../SAMPLE_mapped_and_unmapped.sam` file within
`data_updated/03_Bowtie2_host_filter/#_host-genome/`

The most percent alignments were 0.05-1% alignment, though some rose up to 10% alignments from the chichorium group


#### 3.3 SamTools and bedtools modifications

bowtie unassigned files must be re-formatted with samtools to .sam and .bam for assembly with metaspades
```{bash sam_bed_after_bowtie2, eval = FALSE}

#enter right directory
cd /local/workdir/Daphne/Floral_metagenomics/data_updated/03_Bowtie2_host_filter

for group in ./* ; do
  cd $group
  for sample in ./*/; do
    cd $sample
    #Convert file .sam to .bam
    touch SAMPLE_mapped_and_unmapped.bam
    samtools view -bS SAMPLE_mapped_and_unmapped.sam > SAMPLE_mapped_and_unmapped.bam
    #filter required unmapped reads get unmapped pairs (both ends unmapped)
    samtools view -b -f 12 -F 256 SAMPLE_mapped_and_unmapped.bam > SAMPLE_bothReadsUnmapped.bam 
    # split paired-end reads into separated fastq files .._r1 .._r2
    samtools sort -n -m 5G -@ 2 SAMPLE_bothReadsUnmapped.bam -o SAMPLE_bothReadsUnmapped_sorted.bam
    # sort bam file by read name (-n) for bedtools
    samtools fastq -@ 8 SAMPLE_bothReadsUnmapped_sorted.bam -1 SAMPLE_host_removed_R1.fastq.gz -2 SAMPLE_host_removed_R2.fastq.gz -0 /dev/null -s /dev/null -n
    #Convert bam to fastq
    bedtools bamtofastq -i SAMPLE_bothReadsUnmapped_sorted.bam -fq SAMPLE_host_removed_R1.fastq -fq2 SAMPLE_host_removed_R2.fastq
    cd - > /dev/null # return to previous dir
  done
  cd - > /dev/null
done

#I used this to add "_host" suffix to all folders in 03_Bowtie2_host_filter/host"
for file in 10_Linaria_samples/*; do
  mv $file "${file}_Linaria"
done

#This took very long, ~10 minutes for each group, ~30 minutes for Chichorium
```

This outputs 6 new files into
`Floral_metagenomics/data/03_Bowtie2_host_filter/`
`(# host name)/paired...(sample name)/` - both reads mapped and unmapped
(sam -\> bam -\> sorted) - host removed R1, R2 (fastq.gz -\> .fastq)

the mito-host-removed R1 and R2 files will be subsequently used in the
metaspades assembly







### 4. MetaSpAdes assembly

We will assemble the filtered reads into contigs using metaspades
I read through the manual https://ablab.github.io/spades/running.html
and it seems like there's no parameters of use to us

```{bash Spades_loop, eval=FALSE}

export PATH=/programs/SPAdes-4.0.0/bin:$PATH
cd /local/workdir/Daphne/Floral_metagenomics/data_updated/

declare -a array1=(
[0]=1_Chichorium_samples
[1]=8_Solidago_samples
)
declare -a array2=(
[0]=2_Malus_samples
[1]=4_Ranunculus_samples
[2]=7_Alliaria_samples
[3]=10_Linaria_samples
)
declare -a array3=(
[0]=3_Lotus_samples
[1]=5_Trifolium_samples
[2]=6_Matricaria_samples
[3]=9_Lonicera_samples
)

for host_gen in ${array3[@]}; do
  echo $host_gen
  for sample in 03_Bowtie2_host_filter/"$host_gen"/*/; do
    cd $sample
    pwd
    base=$(basename "$sample")
    mkdir "../../../04_metaspades/"$host_gen"/"$base
    spades.py -1 SAMPLE_host_removed_R1.fastq -2 SAMPLE_host_removed_R2.fastq --meta -t 32 -m 200 -o "../../../04_metaspades/"$host_gen"/"$base
    cd - > /dev/null  # return to previous dir
    pwd 
  done 
done

# Note: you have to create the 04_Metaspades subdirectories before running
  # -1 = file with forward reads
  # -2 = file with reverse reads
  # -t = threads sets the number of processors to use we are using 16 bc thats the default
  # -m = memory limit in Gb. SPAdes terminates this if it reaches 200Gb but the default is 250 Gb
  # -o = output directory
```

Outputs according to spades manual https://currentprotocols.onlinelibrary.wiley.com/doi/full/10.1002/cpbi.102:
1. contigs.fasta — resulting contig sequences in FASTA format;
2. scaffolds.fasta — resulting scaffold sequences in FASTA format;
3. assembly_graph.gfa — assembly graph and scaffolds paths in GFA 1.0 format;
4. assembly_graph.fastg — assembly graph in FASTG format;
5. contigs.paths — paths in the assembly graph corresponding to contigs.fasta;
6. scaffolds.paths — paths in the assembly graph corresponding to scaffolds.fasta;
7. spades.log — file with all log messages.



### 5. MetaQuast check

Provides stats on sample quality

TO RUN: modify the variable host_name for which of the 10 flower hosts you will 
use
```{bash MetaQuast, eval = FALSE}

#### 5.1 Metaquast without Acinetobacter references
I've decided to Run MetaQUAST again, but without using any references. I'll Use a database, like SILVA instead.

To run this faster, I am splitting the 10 groups into 4 larger arrays, and tmux automating this process in 3 different groups

```{bash metaQuast_no_reference, eval = FALSE}

declare -a array1=(
[0]=1_Chichorium_samples

)
declare -a array2=(
[0]=2_Malus_samples
[1]=4_Ranunculus_samples
[2]=7_Alliaria_samples
)
declare -a array3=(
[0]=3_Lotus_samples
[1]=5_Trifolium_samples
)

declare -a array4=(
[0]=6_Matricaria_samples
[1]=8_Solidago_samples
[2]=9_Lonicera_samples
[3]=10_Linaria_samples
)

cd /local/workdir/Daphne/Floral_metagenomics/data_updated/
export PYTHONPATH=/programs/quast-5.2.0/lib64/python3.9/site-packages:/programs/quast-5.2.0/lib/python3.9/site-packages
export PATH=/programs/quast-5.2.0/bin:$PATH

for host_name in ${array5[@]}; do
  echo $host_name
  for sample in 04_metaspades"/$host_name/"*; do
    cd $sample
    pwd
    name="$(basename "$sample" | cut -d'_' -f1-4)" 
    echo $name
    metaquast.py -o ../../../05_metaQUAST/"$host_name/""$name"_metaQuast contigs.fasta
    cd - > /dev/null  # return to previous dir
    pwd 
  done 
done

```
trifolium took longer than others, what kinds of samples were in this?
The output is 
`Floral_metagenomics/data/05_metaQUAST_NO_reference/"host_group"/"sample"/*`
These outputs only show number of contigs (`report.html` and legth stats
`basic_stats/`), and contig lengths 

#### 5.1 MetaQuast with acineto references
Ultimately, I want to see how the number of acineto contigs in data_updated compare to the number of acineto contigs in data/ 

TO RUN: modify the variable host_name for which of the 10 flower hosts you will 
use
```{bash MetaQuast, eval = FALSE}

export PYTHONPATH=/programs/quast-5.2.0/lib64/python3.9/site-packages:/programs/quast-5.2.0/lib/python3.9/site-packages
export PATH=/programs/quast-5.2.0/bin:$PATH

cd /local/workdir/Daphne/Floral_metagenomics/data_updated

#This list contains paths to all acineto reference genomes, from the 04_metaspades sample's directory
list="../../../../data/05_Quast_reference_acineto_genomes/A39.fna,../../../../data/05_Quast_reference_acineto_genomes/ANC4215.fna,../../../../data/05_Quast_reference_acineto_genomes/ANC4422.fna,../../../../data/05_Quast_reference_acineto_genomes/B10A.fna,../../../../data/05_Quast_reference_acineto_genomes/B5B.fna,../../../../data/05_Quast_reference_acineto_genomes/BB226.fna,../../../../data/05_Quast_reference_acineto_genomes/BB362.fna,../../../../data/05_Quast_reference_acineto_genomes/BRTC1.fna,../../../../data/05_Quast_reference_acineto_genomes/CIP110305.fna,../../../../data/05_Quast_reference_acineto_genomes/CIP110357.fna,../../../../data/05_Quast_reference_acineto_genomes/CIP110549.fna,../../../../data/05_Quast_reference_acineto_genomes/DSM14961.fna,../../../../data/05_Quast_reference_acineto_genomes/DSM14964.fna,../../../../data/05_Quast_reference_acineto_genomes/EC031.fna,../../../../data/05_Quast_reference_acineto_genomes/EC034.fna,../../../../data/05_Quast_reference_acineto_genomes/EC115.fna,../../../../data/05_Quast_reference_acineto_genomes/EC24.fna,../../../../data/05_Quast_reference_acineto_genomes/FNA11.fna,../../../../data/05_Quast_reference_acineto_genomes/FNA3.fna,../../../../data/05_Quast_reference_acineto_genomes/NCTC5866.fna,../../../../data/05_Quast_reference_acineto_genomes/NIPH706.fna,../../../../data/05_Quast_reference_acineto_genomes/NIPH991.fna,../../../../data/05_Quast_reference_acineto_genomes/SCC474reference.fna,../../../../data/05_Quast_reference_acineto_genomes/SCC477.fna,../../../../data/05_Quast_reference_acineto_genomes/SM1.fna"

declare -a host_array=(
[0]=1_Chichorium_samples    
[1]=2_Malus_samples
[2]=3_Lotus_samples
[3]=4_Ranunculus_samples
[4]=5_Trifolium_samples
[5]=6_Matricaria_samples
[6]=7_Alliaria_samples
[7]=8_Solidago_samples
[8]=9_Lonicera_samples
[9]=10_Linaria_samples
)



# MODIFY this host_name for the samples you're running
for host_name in ${host_array[@]}; do
  for sample in 04_metaspades"/$host_name/"*; do
    #loop through samples in a flower-host group in metaspades folder
    cd $sample
    name="$(basename "$sample")" 
    echo $name
    # run metaquast, output to 05_metaQUAST, and use $list variable as reference
    metaquast.py -o ../../../05_metaQUAST/05_metaQUAST_acineto/"$host_name/""$name"    contigs.fasta -r "$list" 
    cd - > /dev/null  # return to previous dir
    pwd 
  done
done 
```
The output is 
`Floral_metagenomics/data/05_metaquast/"host_group"/"sample"/*`






### 6. MetaBat2 and preparation

First, rename and group all of the fasta.contigs from the metaspades step into 
one folder, `06_sample_contigs`. This is so that there's one directory 
`06_sample_contigs` with all assembled contigs that contains the host 
(dandelion, apple, etc) in the file name, instead of branched into 
subdirectories by host name. 


#### 6.1 renaming metaspades fasta.contigs into a bwa mem index
```{bash reorganize_sample_contigs}

cd /local/workdir/Daphne/Floral_metagenomics/data_updated/

declare -a host_array=(
[0]=1_Chichorium_samples    
[1]=2_Malus_samples
[2]=3_Lotus_samples
[3]=4_Ranunculus_samples
[4]=5_Trifolium_samples
[5]=6_Matricaria_samples
[6]=7_Alliaria_samples
[7]=8_Solidago_samples
[8]=9_Lonicera_samples
[9]=10_Linaria_samples
)

# rename and group all fasta.contigs from 04_metaspades into 06_sample_contigs
# AND make the bwa index
for host_name in ${host_array[@]}; do
  echo $host_name
  for sample in 04_metaspades"/$host_name/"*; do
    cd $sample
    name="$(basename "$sample" | cut -d'_' -f1-4)" 
    cp ./contigs.fasta ../../../06_sample_contigs/"$name"_"$host_name".fasta
    
    # Do the bwa index
    bwa index ../../../06_sample_contigs/"$name"_"$host_name".fasta
    cd - > /dev/null  # return to previous dir
  done 
done
```
outputs are in `Floral_metagenomics/data_updated/06_sample_contigs`
This takes around 1.5 hours


After the  metaspade fasta files have been renamed, we need to prepare the depth file, but it needs to be first reformatted with sam and bam. 
.sam -> .bam -> depth.txt

#### 6.2 sam bam, depth files
```{bash sam, eval = FALSE}

cd /local/workdir/Daphne/Floral_metagenomics/data_updated

# Make the .sam files in 06_depth_files using bwa mem
for host in ../data/02_trim_unzipped_and_host-grouped/*; do 
  for R1 in $host/*_R1.fastq; do
  
    R2="${R1%%_R1.fastq}_R2.fastq"

    #sample name is the metaspades contigs.fasta file 
    #(this variable doesn't include the host-name suffix in 06_sample_contigs/ files)
    sample_name="$(basename "${R1%%_R1.fastq}" | cut -d'_' -f1-4)"
    output_name="$(basename "$(ls 06_sample_contigs/${sample_name}*.fasta | head -n 1)" | sed 's/\.fasta.*//' )"
    
    #run bwa mem to get the .sam files. inputs are the indices from 06_sample_contigs, and R1/R2 from 02_trimmed_unzipped_host-grouped. output is 06_depth_files
    bwa mem -t 16 "./06_sample_contigs/${output_name}.fasta" $R1 $R2 > 06_depth_files/${output_name}.sam
  done
done

```
output format is  `paired_13702_32540_179738_1_Chichorium.sam` in 
`data_updated/06_depth_files/`

This took 2 hours

```{bash bam, eval = FALSE}

cd /local/workdir/Daphne/Floral_metagenomics/data_updated/06_depth_files

for file in *.sam; do 
  samtools sort $file -o ${file/%.sam/.bam};
done

# sort - samtools sort: sorts the SAM file by coordinate (aka genomic position)
# -o {..}: output filename is the same, but with .bam replacing .sam

```
output format is  `paired_13702_32540_179738_1_Chichorium.bam` in 
`data/06_depth_files/`
This took 1 hour?


This creates the depth files from the .bam files
```{bash depth, eval = FALSE}

cd /local/workdir/Daphne/Floral_metagenomics/data_updated/06_depth_files


# singularity exec: we're running the jgi_summarize inside the container, according to Cornell bioHPC

for file in *.bam; do 
  singularity exec --bind $PWD --pwd $PWD /programs/metabat-2.16/metabat.sif \
    jgi_summarize_bam_contig_depths --outputDepth ${file/%.bam/_depth.txt} $file;
done
# jgi: making a depth file (metabat2 package)
# outputDepth: output filename is same as .sam and .bam, but .bam is replaced with _depth.txt

```

This last command is the first step of metabat2 binning, creating the depth file. 
Metabat manual can be found here: https://gensoft.pasteur.fr/docs/MetaBAT/2.15/

This took 15 min?


#### 6.3 MetaBat2

Here in metabat2, I am running it on the renamed metaspades assembled contigs from 06_sample_contigs, using depth files from 06_depth_files and outputting to 06_MetaBat2
```{bash metabat2, eval = FALSE}

cd /local/workdir/Daphne/Floral_metagenomics/data_updated

for sample in 06_sample_contigs/*.fasta; do 
  depth=$(basename "${sample%%.fasta}")
  echo $depth
  singularity run --bind $PWD --pwd $PWD /programs/metabat-2.16/metabat.sif metabat2 -i $sample -a 06_depth_files/${depth}_depth.txt -o 06_MetaBat2/Metabat2_default/${depth}_bin --seed 42
done

# -i = input contigs from metaspades (necessary)
# -a = depth file (necessary)
# -o = output folder, called 06_MetaBat2 (necessary)


#### WITH BETTER PARAMETERS ####
for sample in 06_sample_contigs/*.fasta; do 
  depth=$(basename "${sample%%.fasta}")
  echo $depth
  singularity run --bind $PWD --pwd $PWD /programs/metabat-2.16/metabat.sif metabat2 -i $sample -a 06_depth_files/${depth}_depth.txt -o 06_MetaBat2/Metabat2_parameters/${depth}_bin --seed 42 -m 1500 -s 100000 --maxEdge 250
done

#Both of these take 2-4 minutes

```

The reasoning for choosing these parameters is on benchling

### 7. CheckM to analyze quality of Metabat2 binning

```{bash checkM2 test, eval = FALSE}

docker1 pull staphb/checkm2

docker1 run --rm \
  -v /local/workdir/Daphne/Floral_metagenomics/databases/checkm2db:/checkm2db \
  -v /local/workdir/Daphne/Floral_metagenomics/data_updated/06_MetaBat2/Metabat2_parameters:/input \
  -v /local/workdir/Daphne/Floral_metagenomics/data_updated/07_CheckM2/CheckM2_parameters:/output \
  staphb/checkm2 checkm2 predict --input /input --output_directory /output -x .fa \
  --database_path /checkm2db/CheckM2_database/uniref100.KO.1.dmnd --threads 30
# this took 2 minutes

```
output is `data_updated/07_CheckM2/checkM2{default or parameter}/`. Here, an output file called `quality_report.tsv` contains the completeness and contamination scores that are of interest


visualize CheckM quality reports in R (use ctrl + enter)
```{r Visualize_CheckM_R, eval = FALSE}
getwd() # should be Floral_metagenomics
checkM2_parameters <- read.table(file="data_updated/07_CheckM2/CheckM2_parameters/quality_report.tsv", sep="\t")

    # change the name (before <-) and subdirectory in 07_CheckM
```



### 8. bins taxonomy with GTDB-k

Gtdb is used to identify taxonomy of the bins produced by MetaBat2. 
Here, we used the Metabat with "parameters" (m1500, s100k, maxE250)

```{bash GTDB-k, eval = FALSE}
# set wd
cd /local/workdir/Daphne/Floral_metagenomics/data_updated

export OMP_NUM_THREADS=16
export PYTHONPATH=/programs/gtdbtk-2.4.0/lib64/python3.9/site-packages:/programs/gtdbtk-2.4.0/lib/python3.9/site-packages
export PATH=/programs/gtdbtk-2.4.0/bin:/programs/hmmer/binaries:/programs/prodigal-2.6.3:/programs/FastTree-2.1.11:/programs/fastANI-1.32:/programs/pplacer-Linux-v1.1.alpha19:/programs/mash-Linux64-v2.3:/programs/skani-0.2.1:$PATH

#This is from ecogenomics.github.io, since the cornell biohpc one is outdated
wget https://data.ace.uq.edu.au/public/gtdb/data/releases/latest/auxillary_files/gtdbtk_package/full_package/gtdbtk_data.tar.gz 
  # this was run originally 6/23
tar xvzf gtdbtk_data.tar.gz -C /local/workdir
  # This took 26 minutes
  
export GTDBTK_DATA_PATH=/workdir/release226

gtdbtk classify_wf --genome_dir 06_MetaBat2/Metabat2_parameters --out_dir 08_gtdbtk_out --cpus 16 --skip_ani_screen --extension fa
# classify_wf = input contigs from metaspades (necessary)
# --genome_dir = bins file (06_MetaBat2/ contains bins)
# --out_dir = outputs to 07_gtdbtk_out
# --cpus 8 = kind of like threads? chatgpt added this
# --skip_ani_screen = use the default database downloaded with wget above
# --extension fa = takes my .fa files, since ftdbtk default is .fasta

#This took 3 hours. Identified 42 genomes as bacterial, 120 markers. Processsed 136677 sequences

```
23 bins of 78 were classifiable, good mix of pseudomonas, pantoea, erwinia, acinetobacter, rosenbergiella, arsenophonus, buchnera

visualize gtdb gtdbtk.bac120.summary.tsv in R (use ctrl + enter)
```{r Visualize_CheckM_R, eval = FALSE}
getwd() # should be Floral_metagenomics
gtdbtk_updated_summary <- read.table(file = "data_updated/08_gtdbtk_out/gtdbtk.bac120.summary.tsv",sep="\t")

gtdbtk_classification <- data.frame(gtdbtk_updated_summary$V1, gtdbtk_updated_summary$V2)


```
### 9. Pseudomonas genome search (BLAST, Bowtie2, MetaQuast)

One possible source of error in the sequencing of our data is index-hopping with pseudomonas samples from Katie that were sequenced at the same time.

To see if this is a problem, I am creating a BLAST database out of Katie's pseudomonas genomes (https://www.ncbi.nlm.nih.gov/bioproject/PRJNA961062) and blast searching our filtered reads or contigs against them. 

I am following NCBI's "Building a BLAST database with your (local) sequences"
https://www.ncbi.nlm.nih.gov/books/NBK569841/

#### 9.1 BLAST
```{bash BLAST_database_and_search, eval = FALSE}

#First, I downloaded all of Katie's .gz files from FTP from the Bioproject linked above, then I unzipped them with for loop and concatenated all into `all_Pseudo_genomes_katie`

cd local/workdir/Daphne/Floral_metagenomics/data_updated/09_blast_of_pseudomonas/Pseudonomas_samples

makeblastdb -in all_Pseudo_genomes_katie.fna -parse_seqids -blastdb_version 5  -out "Pseudomonas_blast_db" -dbtype nucl -title "Katie's Pseudomonas BioProject genomes PRJNA961062"
    #this ran 1.3 second
    #created all_Pseydo_genomes_katie .pdf .phr.pin.pjs.pog ...etc

#inspect info on new database
blastdbcmd -db Pseudomonas_blast_db -info


```

Now, I will try blast searching the samples using these files

#### 9.2 Bowtie2
Instead, Im gonna try aligning with bowtie2: so first I'll build the index, then I'll run bowtie. 
```{bash bowtie2_for_pseudo_search, eval = FALSE}

cd local/workdir/Daphne/Floral_metagenomics/data_updated/09_align_pseudomonas
export PATH=/programs/bowtie2-2.5.1-linux-x86_64:$PATH

#Index for Katie's pseudomonas genomes
bowtie2-build ./Pseudomonas_reference_genomes/all_Pseudo_genomes_katie.fna pseudo_index
    # This took like 2 min


#### Enter reads directory and run bowtie2 ###
cd /local/workdir/Daphne/Floral_metagenomics/data_updated/03_Bowtie2_host_filter
export PATH=/programs/bowtie2-2.5.1-linux-x86_64:$PATH
 
 
declare -a array=(
[0]=10_Linaria_samples
[1]=2_Malus_samples
[2]=3_Lotus_samples
[3]=4_Ranunculus_samples
[4]=5_Trifolium_samples
[5]=6_Matricaria_samples
[6]=7_Alliaria_samples
[7]=8_Solidago_samples
[8]=9_Lonicera_samples
[9]=1_Chichorium_samples
)

for name in ${array[@]}; do
  for sample in /local/workdir/Daphne/Floral_metagenomics/data_updated/03_Bowtie2_host_filter/"$name"/* ; do
    cd $sample
    for R1 in ./*_R1.fastq.gz; do
      base="$(basename "${R1%%_R1.fastq.gz}" | cut -d'_' -f1-4)"
      output_dir="../../../09_align_pseudomonas/${sample}/"
      mkdir -p $output_dir
      echo "$sample" >> ../../../09_align_pseudomonas/pseudo_bowtie2_stdout.txt
      bowtie2 --sensitive-local -p 32 --seed 4 -x ../../../09_align_pseudomonas/pseudo_index -1 $R1 -2 ${R1/R1.fastq.gz/R2.fastq.gz} -S "$output_dir/SAMPLE_mapped_and_unmapped.sam" >> ../../../09_align_pseudomonas/pseudo_bowtie2_stdout.txt 2>&1 
    done     
    cd - > /dev/null
  done   
done

```
The output for bowtie is in `09_align_pseudomonas/pseudo_bowtie2_stdout.txt`, which contains the percent of contamination, and `03_Bowtie2_pseudo_filter/` contains the folders with SAMPLE output files, that have





#### 9.3 MetaQuast
Now I'm gonna try MetaQuast to compare with Bowtie2, and see if it matches teh 10% contamination
```{bash metaquast_pseudo, eval = FALSE}

declare -a array1=(
[0]=1_Chichorium_samples
)
declare -a array2=(
[0]=2_Malus_samples
[1]=4_Ranunculus_samples
[2]=7_Alliaria_samples
)
declare -a array3=(
[0]=3_Lotus_samples
[1]=5_Trifolium_samples
)
declare -a array4=(
[0]=6_Matricaria_samples
[1]=8_Solidago_samples
[2]=9_Lonicera_samples
[3]=10_Linaria_samples
)

cd /local/workdir/Daphne/Floral_metagenomics/data_updated/
export PYTHONPATH=/programs/quast-5.2.0/lib64/python3.9/site-packages:/programs/quast-5.2.0/lib/python3.9/site-packages
export PATH=/programs/quast-5.2.0/bin:$PATH

for host_name in ${array3[@]}; do
  echo $host_name
  for sample in 04_metaspades"/$host_name/"*; do
    cd $sample
    pwd
    name="$(basename "$sample" | cut -d'_' -f1-4)" 
    echo $name
    metaquast.py -o ../../../09_align_pseudomonas/05_metaQUAST_pseudo/"$host_name/""$name"_metaQuast contigs.fasta -r ../../../09_align_pseudomonas/Pseudomonas_reference_genomes/all_Pseudo_genomes_katie.fna
    cd - > /dev/null  # return to previous dir
    pwd 
  done 
done



```



### 10. MetaPhlAn

I'm gonna try to run taxonomy with MetaPhlAn, to compare with Kraken (or instead of kraken if kraken doesn't work). I'm using Katie's data from `` and the metaphlan4 official tutorial https://github.com/biobakery/biobakery/wiki/metaphlan4#1-creating-taxonomic-profiles

```{bash metaphlan setup, eval = FALSE}



##### TEST RUN From Katie's JupyterNotebooks_Cleanedup/2022_NicMeta_Final.ipynb #####
metaphlan 04_metaspades/2_Malus_samples/paired_13702_32540_179775_Malus/scaffolds.fasta --nproc 1 --input_type fasta -o 09_align_pseudomonas/metaphlan/paired_13702_32540_179775_Malus_met.txt  --bowtie2db ../databases/metaphlan


declare -a array1=(
[0]=1_Chichorium_samples
)
declare -a array2=(
[0]=2_Malus_samples
[1]=4_Ranunculus_samples
[2]=7_Alliaria_samples
)
declare -a array3=(
[0]=3_Lotus_samples
[1]=5_Trifolium_samples
)
declare -a array4=(
[0]=6_Matricaria_samples
[1]=8_Solidago_samples
[2]=9_Lonicera_samples
[3]=10_Linaria_samples
)

cd /local/workdir/Daphne/Floral_metagenomics/data_updated
export PYTHONPATH=/programs/metaphlan-4.1.1/lib/python3.9/site-packages:/programs/metaphlan-4.1.1/lib64/python3.9/site-packages
export PATH=/programs/metaphlan-4.1.1/bin:$PATH

for host_name in ${array4[@]}; do
  echo $host_name
  for sample in 04_metaspades"/$host_name/"*; do
    pwd
    base="$(basename "${sample}")"
    scaffolds="$sample/scaffolds.fasta"
    outdir="10_MetaPhlAn"
    metaphlan $scaffolds --nproc 1 --input_type fasta -o ${outdir}/${base}_met.txt  --bowtie2db ../databases/metaphlan
  done 
done
    # --bowtie2db: the database we downloaded, idk why it uses a bowtie db
    # --nproc 5 = number of cores during the mapping stage (tutorial uses 4)
    
  #Generally takes ~7 minutes per sample, so total with 4 tmux took ~2 hours


############### Katie's code, merging and making a heat map ############


cd /local/workdir/Daphne/Floral_metagenomics/data_updated/10_MetaPhlAn


#merge ouputs
/programs/metaphlan-4.1.1/utils/merge_metaphlan_tables.py *_met.txt > merged_abundance_table.txt

# Make a heatmap
/programs/metaphlan-4.0.6/utils/hclust/hclust2.py \
-i merged_abundance_table.txt \
-o metaphlan4_abundance_heatmap_species.png


/programs/metaphlan-4.1.1/utils/hclust2/hclust2.py \
  -i merged_abundance_table.txt \
  -o merged_abundance_table.sqrt_scale_bigger.png \
  --cell_aspect_ratio 1 \
  --flabel_size 3 \
  --metadata_rows 2,3,4 \
  --legend_file HMP.sqrt_scale.legend.png \
  --max_flabel_len 100 \
  --metadata_height 0.075 \
  --slabel_size 3 \
  --max_slabel_len 30 \
  --dpi 300 \
  --image_size 10 \
  --slinkage complete


############# checking number of acinetobacter in 45, 51, 86 output.txt

  awk '/Acinetobacter/ {count++} END {print count}' kraken2_trial_45_output.txt
done

```


### 11. Kraken2 taxonomy

this will use filtered reads, maybe contigs to get taxonomy (better than metaphlan according to a biorxiv paper)

I am using the Kraken2 installation guide to build the databases. https://github.com/DerrickWood/kraken2/wiki/Manual#custom-databases
 Custom databases have 3 steps, as described in the bash chunk below:
 
 https://github.com/DerrickWood/kraken2/blob/master/docs/MANUAL.markdown
```{bash}

#step1: download-taxonomy
#kraken2-build --download-taxonomy --db /local/workdir/Daphne/Floral_metagenomics/databases

#step2: download-library (eg. bacteria, fungi, etc, took like 5 hours for bac, 1 hour for fungi)
#kraken2-build --download-library plant --db  /local/workdir/Daphne/Floral_metagenomics/databases/kraken --threads 64

#step3: build-build ( took 12 hours)
#kraken2-build --build --threads 64 --db /local/workdir/Daphne/Floral_metagenomics/databases/kraken


cd /local/workdir/Daphne/Floral_metagenomics/data_updated
export PATH=/programs/kraken2.1.3:$PATH
export OMP_NUM_THREADS=64 


declare -a array1=(
[0]=1_Chichorium_samples
)
declare -a array2=(
[0]=2_Malus_samples
[1]=4_Ranunculus_samples
[2]=7_Alliaria_samples
)
declare -a array3=(
[0]=3_Lotus_samples
[1]=5_Trifolium_samples
)
declare -a array4=(
[0]=6_Matricaria_samples
[1]=8_Solidago_samples
[2]=9_Lonicera_samples
[3]=10_Linaria_samples
)

for host_name in ${array4[@]}; do
  echo $host_name
  for sample in 03_Bowtie2_host_filter"/$host_name/"*; do
    for paired in ./$sample; do
      R1="${sample}/SAMPLE_host_removed_R1.fastq.gz"
      R2="${sample}/SAMPLE_host_removed_R2.fastq.gz"
      name=$(basename "$paired" | sed 's/^paired_13702_32540_1797//')
      kraken2 --threads 64 --db ../databases/kraken --report 11_Kraken/kraken_${name}_report.txt --paired $R1 $R2
      echo "$name is done"
    done
  done
done

#ran 4:55
# takes around 2 min?

```







### 12. Snippy phylo tree


First, I need the acineto contigs: those I can get per sample in 05_metaQUAST_acineto/{sample}/paired_{num}/combined_reference/contigs_reports/alignments_contigs.tsv

I did cp to copy them to 12_snippy

#### 12.1 isolating acinetobacter contigs
```{bash}

##input: contig.tsv, created in excel by deleting non-pollinis lines from 05_acineto/combined_ref/alignments.contigs.tsv
## getting rid of redundant contig names since contigs in metaquast aligned to multiple acineto's, just output unique NODEs from contigs.fasta
cd /local/workdir/Daphne/Floral_metagenomics/data_updated/12_snippy/snippy_input_acineto_contig_files

awk '
  NR == FNR {
    # Read and collect all NODE IDs from the contig list (1st pass)
    for (i = 1; i <= NF; i++) {
      if ($i ~ /^NODE/) {
        if (!seen[$i]++) {
          node_ids[$i]
        }
      }
    }
    next
  }

  # Process the FASTA file (2nd pass)
  /^>/ {
    keep = 0
    if (substr($1, 2) in node_ids) {
      keep = 1
    }
  }

  keep
' FS='\t' OFS='\t' 93_contig.tsv ../../04_metaspades/10_Linaria_samples/paired_13702_32540_179793_Linaria/contigs.fasta > 93_acineto_contigs.fasta

```
output is called `XX_acineto_contigs.fasta` in `data_updated/12_snippy/12_acineto_contig_files/`


Now, we can use these fasta files with acinetobacter contigs to run snippy
steps: 
snippy-core
snippy-clean_full_aln
snp-sites
FastTree

#### 12.2 Snippy Pollinis
```{bash, eval = FALSE}

export PATH=/programs/snippy-4.6.0/bin:/programs/snippy-4.6.0/binaries/linux:$PATH
export PATH=/programs/seqtk:$PATH
export PATH=/programs/samclip:$PATH
export PATH=/programs/vt:$PATH

cd /local/workdir/Daphne/Floral_metagenomics/data_updated/12_snippy

#1. run snippy of 1 sample (44 or 45 or 48 or 86) into output snippy
snippy  --outdir snippy_outdir/sample_44 --ref A-pollinis_FNA11.fna --ctgs snippy_input_acineto_contig_files/93_acineto_contigs.fasta --force

#2. do snippy-core on all samples
SAMPLES_pol="../../snippy_outdir/sample_44
../../snippy_outdir/sample_45
../../snippy_outdir/sample_48
../../snippy_outdir/sample_86"
cd snippy_core/pollinis_core
snippy-core --ref ../../A-pollinis_FNA11.fna $SAMPLES_pol

#3. make clean.full.aln
snippy-clean_full_aln core.full.aln > clean.full.aln

#4. make clean.core.aln with snp-sites
/programs/snp-sites-2.5.1/bin/snp-sites
export PATH=/programs/snp-sites-2.5.1/bin:$PATH
snp-sites -c clean.full.aln > clean.core.aln 

#5. make a phylogenomic SNP tree with FastTree
export PATH=/programs/FastTree-2.1.11:$PATH
FastTree -gtr -nt clean.core.aln > clean.core.tree

#6. make phylogenomic SNP tree with IQ tree
/programs/iqtree-2.2.2.6-Linux/bin/iqtree2 -s clean.core.aln -m MFP 
  #moved to IQ_tree/

```


#### 12.3 Snippy Nectaris

```{bash}
export PATH=/programs/snippy-4.6.0/bin:/programs/snippy-4.6.0/binaries/linux:$PATH
export PATH=/programs/seqtk:$PATH
export PATH=/programs/samclip:$PATH
export PATH=/programs/vt:$PATH

cd /local/workdir/Daphne/Floral_metagenomics/data_updated/12_snippy

#1. run snippy of 1 sample (75,93) into output snippy
snippy  --outdir snippy_outdir/sample_93 --ref A-nectaris_BB226.fna --ctgs snippy_input_acineto_contig_files/93_acineto_contigs.fasta --force

#2. join reads in snippy-core
SAMPLES_nect="../../snippy_outdir/sample_75
../../snippy_outdir/sample_93"
cd snippy_core/nectaris_core
snippy-core --ref ../../A-nectaris_BB226.fna $SAMPLES_nect
  #surprisingly worked with no errors


#3. make clean.full.aln with snippy-clean_full_aln 
snippy-clean_full_aln core.full.aln > clean.full.aln 

#NOT using gubbins because we dont have to
#source /programs/miniconda3/bin/activate gubbins-3.1.2
#run_gubbins.py -p gubbins clean.full.aln


#4. make clean.core.aln with snp-sites
/programs/snp-sites-2.5.1/bin/snp-sites
export PATH=/programs/snp-sites-2.5.1/bin:$PATH
snp-sites -p -c clean.full.aln > clean.core.phy
  #can output a phylip file with -p for IQ tree

#5. make a phylogenomic SNP tree with FastTree
export PATH=/programs/FastTree-2.1.11:$PATH
FastTree -gtr -nt clean.core.aln > clean.core.tree
  
#6. make phylogenomic SNP tree with IQ tree
/programs/iqtree-2.2.2.6-Linux/bin/iqtree2 -s clean.core.phy -m MFP
   #moved to IQ_tree/ 
 

```




### 13. Kaiju taxonomy Reads 
  
  this is an alternative to Kraken, since it seems to give a lot of false positives 

#### 13.0 download kaiju database
```{bash kaiju, eval = FALSE}
  
#the kaiju download doesn't work, because one GFC: GCF_050881545.1 doesn't download and outputs error 404.
#Instead, I am downloading kaiju_db_nr (81 GB) database via getwd of this site I found online:
https://bioinformatics-centre.github.io/kaiju/downloads.html 

wget https://kaiju-idx.s3.eu-central-1.amazonaws.com/2024/kaiju_db_refseq_2024-08-25.tgz  
 
#then extract  
tar xzf kaiju_db_nr_2024-08-25.tgz
 
## test to see if downloads worked:
kaiju -t nodes.dmp -f kaiju_db_refseq.fmi -i pectin_lyase_A-apis_test.fasta -o test.out 
  #outputted to test.out in 3.41 min
 
```

#### 13.1 run kaiju on all the files
NOTE: if multiple tmux are running kaiju, it will crash. just run one
```{bash tmux_kaiju, eval = FALSE}

declare -a array1=(
[0]=9_Lonicera_samples
[1]=10_Linaria_samples
[2]=4_Ranunculus_samples
[3]=3_Lotus_samples
[4]=5_Trifolium_samples
[5]=8_Solidago_samples
[6]=7_Alliaria_samples
[7]=6_Matricaria_samples
[8]=1_Chichorium_samples
[9]=2_Malus_samples
)

declare -a array2=(
[0]=2_Malus_samples
[1]=1_Chichorium_samples
)

cd /local/workdir/Daphne/Floral_metagenomics/data_updated/03_Bowtie2_host_filter 

#This runs kaiju on all samples from 03_bowtie
for host_name in ${array2[@]}; do
  for R1 in "$host_name/"*"/SAMPLE_host_removed_R1.fastq"; do
    export PATH=/programs/kaiju-1.10.0/bin:$PATH
    echo $(date)
    R2="${R1/R1.fastq.gz/R2.fastq}"
    output_name=$(echo "$R1" | cut -d'/' -f2)
    kaiju_path="/local/workdir/Daphne/Floral_metagenomics/databases/kaiju_nr_euk"
    kaiju -t "${kaiju_path}/nodes.dmp" -f "${kaiju_path}/kaiju_db_nr_euk.fmi" -i $R1 -j $R2 -o ../13_Kaiju/1-1_raw_kaiju_euk/${output_name}_kaiju.txt -z 30
    echo $(date)
  done
done

#this creates all_classified_kaiju_stats.out, containing number of classified reads, total reads, and the percent, for each sample
cd /local/workdir/Daphne/Floral_metagenomics/data_updated/13_Kaiju/1-1_raw_kaiju_euk
#check percent of lines that are "classified vs unclassified"
for file in ./*.txt; do
  echo $file >> all_classified_kaiju_euk_stats.out
   #num files with C, total files, C's percent of total
  awk '{total++} $1 == "C" {c++} END {print c, total, c/total}'  $file >> all_classified_kaiju_euk_stats.out
done
```
Outputs for both of these scripts are saved in /13_Kaiju/1_raw_kaiju

#### 13.2 assign genus taxonomy to kaiju reads
Once kaiju files are made, isolate the acinetobacter files
```{bash}

#delete line that contains taxon 3242193 and 3242193 bc they cause errors for kaiju assign taxon
for file in ./paired*; do
  sed -i '/3242192/d' $file
  sed -i '/3242193/d' $file
done
  
# Run kaiju-addTaxonNames -r genus to add genus taxonomy to all reads  
#NOTE: remember that I have the right path for the kaiju database I want to use
kaiju_path="/local/workdir/Daphne/Floral_metagenomics/databases/kaiju_refseq"
for files in ./*.txt; do
  name=$(basename "$files" | sed 's/^paired_13702_32540_1797//;s/_kaiju.txt$/_kaiju_genus.out/')
  kaiju-addTaxonNames -t ${kaiju_path}/nodes.dmp -n      ${kaiju_path}/names.dmp -i $files -r genus -o ../2_kaiju_taxonomy/$name
done
  
#print to acineto_stats.out number of "Acinetobacter" per sample
for genus_files in ./*_genus.out; do
  echo $(basename "$genus_files") >> acineto_stats.out
  awk '/Acinetobacter/ {count++} END {print count}' $genus_files >> acineto_stats.out
done
```
Outputs for these scripts are saved in /13_Kaiju/2_kaiju_taxonomy

#### 13.3 obtain read names for all acinetobacter reads
```{bash}

# extract reads with 'genus of interest'
#separating by tabs /t, if 4th column = "acineto", print 2nd column (read name)
for file in 2_kaiju_taxonomy/*Lonicera_kaiju_genus.out; do
  output=$(basename $file | cut -d'_' -f1-3)
awk -F"\t" '$4 == "Acinetobacter; " {print $2}' $file > 3_acineto_read_names/${output}_acineto_read_name.txt
done

# sort them into host name folders, to make it easier to match with 03_bowtie2_filtered_reads 
```
outputs are in 13_Kaiju/3_acineto_read_names/{host name}

#### 13.4 Extract reads from 03_bowtie2 files using our kaiju name files
tip: change host_name variable at the top of this chunk then just press "run"
```{bash}

host_name="10_Linaria"

export PATH=/programs/seqtk:$PATH
cd /local/workdir/Daphne/Floral_metagenomics/data_updated/13_Kaiju


for name_file in 3_acineto_read_names/${host_name}_samples/*; do
  prefix=$(basename "$name_file"); prefix=${prefix%%_kaiju_acineto_read_name.txt}
  echo $prefix ######################################
  
  # Loop through 03_bowtie/subdirectories
  for dir in ../03_Bowtie2_host_filter/${host_name}_samples/paired_13702_32540_1797*; do
      # Remove everything before first underscore
      suffix=$(echo "$dir" | awk -F'13702_32540_1797' '{print $2}') 
      
      # if acineto_name_list and bowtie reads name match, do seqtk
      if [[ "$prefix" == "$suffix" ]]; then
        
        #create temp name files with added /1 and /2
        sed 's/$/\/1/' $name_file > tmp_R1.txt
        sed 's/$/\/2/' $name_file > tmp_R2.txt
        
        #use seqtk to extract reads witha name file
        seqtk subseq $dir/SAMPLE_host_removed_R1.fastq tmp_R1.txt > 4_acineto_reads/${prefix}_R1.fastq
        seqtk subseq $dir/SAMPLE_host_removed_R2.fastq tmp_R2.txt > 4_acineto_reads/${prefix}_R2.fastq
        
        # remove temp files
        rm tmp_R1.txt tmp_R2.txt
        
      fi
  done
done
```
outputs are in 13_Kaiju/4_acineto_reads

#### 13.5 making a heatmap

1. get genus-level taxonomy and abundances with kaiju2table (bash)
2. modify the table to be a matrix of taxa vs sample (R)
3. generate a heatmap

In this section, I alter the code "kaiju_files", "kaiju_path" and the output of 
kaiju2table DIRECTLY in the script box to make varied heatmaps using different 
kaju databases, taxon levels, etc 
```{bash}
cd /local/workdir/Daphne/Floral_metagenomics/data_updated/13_Kaiju/1-1_raw_kaiju_euk
export PATH=/programs/kaiju-1.10.0/bin:$PATH
# 1. add taxonomy to all samples with kaiju2table
# Run kaiju-addTaxonNames -r genus to add genus taxonomy to all reads  
#updated path for what kaiju db you want to use 
kaiju_path="/local/workdir/Daphne/Floral_metagenomics/databases/kaiju_nr_euk"

kaiju_files="
paired_13702_32540_179738_Chichorium_kaiju.txt
paired_13702_32540_179739_Solidago_kaiju.txt
paired_13702_32540_179740_Chichorium_kaiju.txt
paired_13702_32540_179741_Solidago_kaiju.txt
paired_13702_32540_179742_Ranunculus_kaiju.txt
paired_13702_32540_179743_Malus_kaiju.txt
paired_13702_32540_179744_Malus_kaiju.txt
paired_13702_32540_179745_Trifolium_kaiju.txt
paired_13702_32540_179746_Solidago_kaiju.txt
paired_13702_32540_179747_Trifolium_kaiju.txt
paired_13702_32540_179748_Chichorium_kaiju.txt
paired_13702_32540_179749_Malus_kaiju.txt
paired_13702_32540_179750_Lotus_kaiju.txt
paired_13702_32540_179751_Matricaria_kaiju.txt
paired_13702_32540_179752_Chichorium_kaiju.txt
paired_13702_32540_179753_Lotus_kaiju.txt
paired_13702_32540_179754_Lonicera_kaiju.txt
paired_13702_32540_179755_Alliaria_kaiju.txt
paired_13702_32540_179756_Ranunculus_kaiju.txt
paired_13702_32540_179757_Trifolium_kaiju.txt
paired_13702_32540_179758_Lotus_kaiju.txt
paired_13702_32540_179759_Lotus_kaiju.txt
paired_13702_32540_179760_Chichorium_kaiju.txt
paired_13702_32540_179761_Malus_kaiju.txt
paired_13702_32540_179762_Malus_kaiju.txt
paired_13702_32540_179763_Chichorium_kaiju.txt
paired_13702_32540_179764_Chichorium_kaiju.txt
paired_13702_32540_179765_Alliaria_kaiju.txt
paired_13702_32540_179766_Malus_kaiju.txt
paired_13702_32540_179767_Lonicera_kaiju.txt
paired_13702_32540_179768_Chichorium_kaiju.txt
paired_13702_32540_179769_Chichorium_kaiju.txt
paired_13702_32540_179770_Chichorium_kaiju.txt
paired_13702_32540_179771_Chichorium_kaiju.txt
paired_13702_32540_179772_Chichorium_kaiju.txt
paired_13702_32540_179773_Chichorium_kaiju.txt
paired_13702_32540_179774_Chichorium_kaiju.txt
paired_13702_32540_179775_Malus_kaiju.txt
paired_13702_32540_179776_Solidago_kaiju.txt
paired_13702_32540_179777_Chichorium_kaiju.txt
paired_13702_32540_179778_Chichorium_kaiju.txt
paired_13702_32540_179779_Chichorium_kaiju.txt
paired_13702_32540_179780_Chichorium_kaiju.txt
paired_13702_32540_179781_Chichorium_kaiju.txt
paired_13702_32540_179782_Chichorium_kaiju.txt
paired_13702_32540_179783_Lotus_kaiju.txt
paired_13702_32540_179784_Chichorium_kaiju.txt
paired_13702_32540_179785_Malus_kaiju.txt
paired_13702_32540_179786_Matricaria_kaiju.txt
paired_13702_32540_179787_Chichorium_kaiju.txt
paired_13702_32540_179788_Chichorium_kaiju.txt
paired_13702_32540_179789_Chichorium_kaiju.txt
paired_13702_32540_179790_Chichorium_kaiju.txt
paired_13702_32540_179791_Chichorium_kaiju.txt
paired_13702_32540_179792_Chichorium_kaiju.txt
paired_13702_32540_179793_Linaria_kaiju.txt
"

kaiju2table -t ${kaiju_path}/nodes.dmp -n ${kaiju_path}/names.dmp -r phylum -m 1.0 -u -o ../5_heatmap/kaiju_phylum_summary.tsv $kaiju_files

```


change input for kaiju_df
```{r modify-and-reformat_kaiju2table, eval = FALSE}
#setup
setwd("local/workdir/Daphne/Floral_metagenomics/data_updated")
library(tidyverse)

# Read the Kaiju summary file
kaiju_df <- read_tsv("13_Kaiju/5_heatmap/kaiju_phylum_summary.tsv")

# Clean up sample names (remove path and _kaiju.txt)
kaiju_df <- kaiju_df %>%
  mutate(sample = basename(file) %>%
        str_replace("_kaiju\\.txt$", "")%>%
        str_replace("paired_13702_32540_1797", ""))
        
# Create abundance matrix: taxon_name x sample (values = reads)
abundance_matrix <- kaiju_df %>%
  select(sample, taxon_name, reads) %>%
  group_by(taxon_name, sample) %>%
  summarise(reads = sum(reads), .groups = "drop") %>%
  pivot_wider(names_from = sample, values_from = reads, values_fill = 0)

# Write to CSV
write_csv(abundance_matrix, "13_Kaiju/5_heatmap/kaiju_abundance_matrix.csv")
```

```{r kaiju_heatmap, eval = FALSE}
library(pheatmap)

data <- read.csv("13_Kaiju/5_heatmap/kaiju_abundance_matrix.csv", row.names=1)
pheatmap(data, scale="row", clustering_distance_rows="euclidean",
         clustering_distance_cols="euclidean", color = colorRampPalette(c("purple4", "white", "goldenrod1"))(50), cellwidth = 6,
  cellheight = 6, fontsize = 5)
```



### 14 Snippy again with Kaiju reads


#### 14.1 create tab input files for snippy-multi
snippy multi allows you to check multiple isolates against same reference, using names file with tabs 
```{bash}

# input of all samples (minus refernce genomes)
name="all_samples"
cd /local/workdir/Daphne/Floral_metagenomics/data_updated
printf "" > 14_snippy_kaiju/inputs_for_snippy-multi/$name.tab #clear output
for R1 in 13_Kaiju/4_acineto_reads/*R1.fastq; do
  ID="$(basename "${R1%%_R1.fastq}")"
  R2=${R1/R1.fastq/R2.fastq}
  printf "%s\t%s\t%s\n" "14_snippy_kaiju/snippy/snippy-multi_$name/$ID" "$R1" "$R2" >> 14_snippy_kaiju/inputs_for_snippy-multi/$name.tab
done


# input of all reference genomes
name_genome="ref_nectaris"
printf "" > 14_snippy_kaiju/inputs_for_snippy-multi/$name_genome.tab 
for genome in 13_Kaiju/RAST_nectar_genomes/A-nectaris*.fna; do
  ID="$(basename "${genome%%.fna}")"
  printf "%s\t%s\n" "14_snippy_kaiju/snippy/snippy-multi_$name_genome/$ID" "$genome">> 14_snippy_kaiju/inputs_for_snippy-multi/$name_genome.tab
done

  # note: these input files have paths from /local/workdir/Daphne/Floral_metagenomics/data_updated, so make sure future snippy steps also are based from this path

```



Now, we can use these fasta files with acinetobacter contigs to run snippy
steps: 
snippy
snippy-core
snippy-clean_full_aln
snp-sites
FastTree

#### 14.2 Snippy-multi on samples and reference genomes

This does snippy core so I think I should delete files that have no data (samples 46 and 83)
```{bash}

export PATH=/programs/snippy-4.6.0/bin:/programs/snippy-4.6.0/binaries/linux:$PATH
export PATH=/programs/seqtk:$PATH
export PATH=/programs/samclip:$PATH
export PATH=/programs/vt:$PATH

cd /local/workdir/Daphne/Floral_metagenomics/data_updated

#snippy multi for all samples (minus refernce genomes)
name="allsample_nectaris"
snippy-multi 14_snippy_kaiju/inputs_for_snippy-multi/$name.tab --ref 13_Kaiju/RAST_nectar_genomes/A-nectaris-BB226.fna --cpus 20 --mincov 4 > 14_snippy_kaiju/snippy/snippy-multi_$name.sh
#changed snippy-core ref to be FNA11 or BB226
#add prefix for output --prefix '14_snippy_kaiju/snippy/snippy-multi_all_samples/core' 

sh ./14_snippy_kaiju/snippy/snippy-multi_$name.sh #run snippy


#snippy multi for all reference genomes
name_genome="ref_nectaris"
snippy-multi 14_snippy_kaiju/inputs_for_snippy-multi/$name_genome.tab --ref 13_Kaiju/RAST_nectar_genomes/A-nectaris-BB226.fna --cpus 20 > 14_snippy_kaiju/snippy/snippy-multi_$name_genome.sh
#changed snippy-core ref to be FNA11
#add prefix for output --prefix '14_snippy_kaiju/snippy/snippy-multi_ref_nectaris/core'

sh ./14_snippy_kaiju/snippy/snippy-multi_$name_genome.sh # run snippy


#snippy core for top 6 genomes and all reference genomes
sh ./14_snippy_kaiju/snippy/snippy_core-top6_all_genomes.sh
 
```




#### 14.3 from snippy to IQ tree
```{bash, eval = FALSE}
#Remember to cd into 14_snippy_kaiju/snippy/{your working folder}

#3. make clean.full.aln
snippy-clean_full_aln core-39.full.aln > 39-clean.full.aln

#4. make clean.core.aln with snp-sites
/programs/snp-sites-2.5.1/bin/snp-sites
export PATH=/programs/snp-sites-2.5.1/bin:$PATH
snp-sites -c 39-clean.full.aln > 39-clean.core.aln 

#5. make phylogenomic SNP tree with IQ tree
/programs/iqtree-2.2.2.6-Linux/bin/iqtree2 -s 39-clean.core.aln -m MFP 
  

```


### CURL
```{bash}

ftp://ftp.mg-rast.org/tools/upload/mg-inbox.py

python mg-inbox.py --help
# from chatgpt
curl -H "auth: <YOUR_API_KEY>" -F "upload=@your_sample.fastq.gz" "https://api.mg-rast.org/1/inbox"

cd /local/workdir/Daphne/Floral_metagenomics/databases/MG-RAST-Tools
#from mg rast website:
curl -X POST -H "auth: aN46PrhE8kpzsBX23AYdQKFXT" -F "upload=@/paired_13702_32540_179793_Lonicera/contigs.fasta"  

curl -X POST \
     -H "auth: aN46PrhE8kpzsBX23AYdQKFXT" \
     -F "upload=@<path/to/yourfile.fastq>" \
     "https://api.mg-rast.org/inbox"

# my webkey for 8/4/2025 is:aN46PrhE8kpzsBX23AYdQKFXT (should work
# like an API)

#chatgpt again
curl -H "auth: aN46PrhE8kpzsBX23AYdQKFXT" -F "upload=@../../data_updated/03_Bowtie2_host_filter/5_Trifolium_samples/paired_13702_32540_179745_Trifolium/SAMPLE_host_removed_R1.fastq" "https://api.mg-rast.org/1/inbox"


#this has the raw read sitting in data_updated
curl -X POST -H "auth: aN46PrhE8kpzsBX23AYdQKFXT" -F "upload=@13702_32540_179745_H5MH5AFX5_metagenomic_mix_H06_FN128_TAGGCATG_TCGCATAA_R1.fastq.gz" "https://api.mg-rast.org/1/inbox"


echo -e "@SEQ_ID\nGATTACA\n+\nIIIIIII" > test.fastq
gzip test.fastq
curl -X POST -H "auth: aN46PrhE8kpzsBX23AYdQKFXT" -F "13702_32540_179745_H5MH5AFX5_metagenomic_mix_H06_FN128_TAGGCATG_TCGCATAA_R1.fastq.gz" "https://api.mg-rast.org/1/inbox"

curl -v -H "auth: aN46PrhE8kpzsBX23AYdQKFXT"  -F "upload=@13702_32540_179745_H5MH5AFX5_metagenomic_mix_H06_FN128_TAGGCATG_TCGCATAA_R1.fastq.gz" "https://api.mg-rast.org/1/inbox" > out 2>error

#git link from chatgpt:
https://github.com/MG-RAST/MG-RAST-Upload

```

